### 2024-12-02 20240201_T-reX_JournalUploadFolder/main.tex
always one more...

%% 
%% Copyright 2019-2024 Elsevier Ltd
%% 
%% This file is part of the 'CAS Bundle'.
%% --------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.3c of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.3c or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%%
%
% T-reX manuscript using the elsevier template for JRC
%
% Repository for the T-reX manuscript at https://github.com/Stew-McD/T-reX_Publication
% Repository for the T-reX python program at https://github.com/Stew-McD/T-reX
%
% S.C.McDowall - github/Stew-McD
%
%%%%

@@

%\usepackage[numbers]{natbib}
\usepackage[authoryear]{natbib}
%\usepackage[authoryear,longnamesfirst]{natbib}

%Author macros
\def\tsc#1{\csdef{#1}{\textsc{\lowercase{#1}}\xspace}}
\tsc{SCM}
\tsc{EL}
\tsc{SC}
\tsc{CFB}
%%%

% Essential packages
\usepackage{graphicx}
\usepackage{float} % For precise figure placement
\usepackage[authoryear]{natbib} % Use natbib for citations

% Optional packages (uncomment only if necessary)
%\usepackage{subcaption} % For subfigures
%\usepackage{xcolor} % If custom colours are needed
%\usepackage{lineno} % For line numbering (use \linenumbers to activate)

% Ensure compatibility
\usepackage{booktabs} % For professional-quality tables

% \setlength{\linenumbersep}{5pt}
%
\usepackage{graphicx}

% ------------------------------------------------------

\begin{document}

\let\WriteBookmarks\relax
\def\floatpagepagefraction{1}
\def\textpagefraction{.001}

@@

% Short author
\shortauthors{S2. McDowall et~al.}

% Main title of the paper
\title[mode = title]{T-reX: Quantifying waste and material footprints in current and future Life Cycle Assessment (LCA) databases}

% Authors
\author[1]{Stewart Charles McDowall}[orcid=0000-0002-5688-1279]
\ead{s.c.mcdowall@cml.leidenuniv.nl}
\cormark[1]
\cortext[cor1]{Corresponding author:}
\credit{Methodology, Software, Validation, Formal analysis, Investigation, Data curation, Writing - original draft, Writing - review \& editing, Visualization}

@@

\affiliation[1]{
	organization={Institute of Environmental Sciences (CML), Leiden University},
	addressline={P.O. Box 9518},
	city={Leiden},
	postcode={2300 RA},
	state={South Holland},
	country={The Netherlands}}

% Abstract\\
\begin{abstract}
The quintessential principle of the circular economy is to minimise material consumption and waste generation.
Thus, identifying and quantifying waste and material flows is critical.
 Life Cycle Assessment (LCA) is a powerful quantitative method for this,
especially given its capacity to expose the details of an activity's entire life cycle, 
often guiding the implementation of circular principles to where they could be most effective.

Pursuant to these goals the authors have developed `T-reX', an analytical tool
for LCA, written in Python. T-reX extends on the existing computational
ecosystem centered on Brightway and Premise and can facilitate the
quantification of user-defined supply chain demands for activities in current
and prospective scenarios. T-reX streamlines database manipulation for LCA
practitioners, integrating methods to aggregate and analyse demand inventories
that could expose hotspots of hidden risk.

With a simple case study using Lithium-ion batteries, we demonstrate some of
T-reX's potential by detailing and exploring aspects of their waste and
material footprints. Since such footprints are often linked to negative
externalities, T-reX could support sustainable decision-making and contribute
to the development of the circular economy by promoting the understanding of
our material consumption and waste generation.
\end{abstract}

% Graphical abstract
\begin{graphicalabstract}
	\includegraphics[width=\textwidth]{figures/T-reX_graphical-abstract.pdf}
\end{graphicalabstract}

\begin{keyword}
circular economy \sep wste \sep material footprint \sep life cycle assessment \sep critical raw materials \sep supply chain
\end{keyword}

% Highlights
\begin{highlights}
	\item T-reX, a new open-source program for quantifying waste and material flows in	Life Cycle Assessment (LCA)
	\item Assesses supply risks by calculating demand inventories for critical materials
	\item Simplifies quantification of user-specified waste and material footprint categories
	\item Rapidly identifies hotspots of waste generation and material demand in supply	chains
	\item Presents a case study of five lithium-ion batteries to demonstrate T-reX's potential
\end{highlights}

% Keywords
\begin{keywords}
	circular economy \sep waste \sep material footprint \sep life cycle assessment \sep critical raw materials \sep supply chain
\end{keywords}

%\maketitle

%\nocite{*}

\begin{table*}
	\centering
	\caption*{List of abbreviations and terminology}\label{tab:abbreviations}
	\begin{tabular}{ll}
		\toprule
		CCS	              & Carbon Capture and Storage                                          \\
		CF                & Carbon Footprint                                                    \\
		EF                & Ecological Footprint                                                \\
		CML               & Centrum voor Milieuwetenschappen (Centre for Environmental Science) \\
		CRM               & Critical Raw Material                                               \\
		CPC               & Cooperative Patent Classification                                   \\
		CSI               & Crustal Scarcity Indicator                                          \\
		CSP               & Crustal Scarcity Potential                                          \\
		EDIP              & Environmental Design of Industrial Products                         \\
		EEIOA             & Environmentally Extended Input-Output Analysis                      \\
		EOL               & End-of-Life                                                         \\
		FOEN              & Swiss Federal Office for the Environment                            \\
		IAM               & Integrated Assessment Model                                         \\
		IE                & Industrial Ecology                                                  \\
		LCA               & Life Cycle Assessment                                               \\
		LCI               & Life Cycle Inventory                                                \\
		LCIA              & Life Cycle Impact Assessment                                        \\
		LFP               & Lithium Iron Phosphate                                              \\
		Li-ion            & Lithium-ion                                                         \\
		LiMn\(_2\)O\(_4\) & Lithium Manganese Oxide                                             \\
		MF                & Material Footprint                                                  \\
		MFA               & Material Flow Analysis                                              \\
		NCA               & Nickel Cobalt Aluminum                                              \\
		NMC               & Nickel Manganese Cobalt                                             \\
		PWF               & Product Waste Footprint                                             \\
		RCP               & Representative Concentration Pathway                                \\
		ReCiPe            & A standard LCIA method set                                          \\
		SDG               & Sustainable Development Goals                                       \\
		SSPs              & Shared Socioeconomic Pathways                                       \\
		T-reX             & The Tool for analysing re-X in LCA                                  \\
		UBP               & Umweltbelastungspünkte (Environmental Impact Points)                \\
		UN                & United Nations                                                      \\
		WF                & Waste Footprint                                                     \\
		\bottomrule
	\end{tabular}
\end{table*}

% Main text

% Introduction
\section{Introduction}\label{sec:introduction}

\subsection{Introduction to T-reX}\label{sec:intro-trex}

The development of a circular economy has become a central focus in the pursuit
of sustainability objectives toward curtailment of our environmental footprint
within planetary boundaries~\citep{eu2019greendeal,	eu2020circ,nl2023ceplan,nl2016ceplan,pardo2018ce,ellenmacarthur2015ce}.

Fundamental to this development is a decrease in primary material consumption
and a reduction of life cycle waste through the implementation of `re-X'
strategies (e.g., refuse, rethink, design for---and implementation of---repair,
remanufacturing and recycling)~\citep{reike2018rex, eu2022ecodesign,
	eu2022repair, eu2015reman}. In addition to circular economy goals, contemporary
geopolitical tensions in an ever more globalised world have highlighted the
vulnerability of many advanced economies to intentional supply
disruptions---wrought as an act of competition or even outright
hostility~\citep{jrc2023supplychain,hartley2024cepolitics,berry2023crm}.

The concept of the `footprint' as an environmental sustainability indicator
began with the Ecological Footprint
(EF)~\citep{wackernagel1994ecologicalfootprint} and after being popularised by
the Carbon Footprint (CF)~\citep{cucek2015environmentalfootprints} the
`footprint family' has adopted many additional metrics---albeit without yet
coalescing into a coherent or consistent
framework~\citep{giampietro2014footprintstonowhere,
	vanham2019footprints,ridoutt2013footprints}. More recently, the footprint
collection has been extended to include the Material Footprint
(MF)~\citep{weidmann2013materialfootprint}, which is often encountered in the
`macro methods' of Industrial Ecology (IE), such as environmentally Extended
Input-Output Analysis (EEIOA)~\citep{lenzen2022materialfootprint} and Material
Flow Analysis (MFA)~\citep{schaffartzik2013mfafootprint}. Whether at the level
of products, entire industries, nations, or even continents, the material
footprint aims to quantify the total material consumed in supply chains. It has
been shown that the MF can be `highly representative of damage to human health
and biodiversity'~\citep{steinmann2017resourcefootprints} and indeed, this
metric was recently beatified by the United Nations (UN), becoming the `core
official indicator' for targets 8.4 and 12.2 of the Sustainable Development
Goals (SDGs)~\citep{lenzen2022materialfootprint}.

@@

WF and MF, the two metrics of focus in the study, can provide a comprehensive
assessment of potential environmental impacts across the supply chain,
encapsulating both resource use and pollution/waste generation, while offering
insights at diverse scales, from individual activities to global systems,
facilitating communication with a diverse range of stakeholders. Thus, to
reduce the negative externalities of human consumption and improve supply chain
resilience, it is essential to uncover, dis-aggregate, and quantify the material
and waste footprints of human activities in as much detail as
possible~\citep{bisinella2024wastelca,
	towa2020wastefootprint,berger2020mineralsinlca,sonderegger2020mineralsinlca}.

@@

To facilitate the quantification of waste and material flows in LCA, we
developed a Python package to extend the \texttt{brightway} open-source LCA
framework~\citep{mutel2017brightway} and track these exchanges by translating
them into inventory indicators and `pseudo' LCA impact (LCIA) categories. T-reX
is `T(ool for) re-X' (reduce, reuse, recycle, etc.\@ \ldots) and enables LCA
practitioners to manipulate their databases to allow them to easily aggregate
the mass and volume of any desired exchange, and to create flexible categories
that differentiate between material categories, waste types, and treatment or
EOL handling.

@@

~\cite{laurenti2023wastefootprint} developed a method to calculate the waste footprint of a product or service based on solving the demand vectors of the activities, while also proposing simple measures to quantify waste `hazardousness' and `circularity' with this data. The method presented, as the authors explain, suffers from errors due to double counting due to the way that exchanges are identified and `tagged' . In T-reX, the source database is deconstructed into a list of separate exchanges, and only those explicitly defined as hazardous are marked as such. The `T-reX' Python software tool presented herein provides a more flexible, transparent, and user-friendly approach to quantifying waste flows in LCA\@. Moreover, the utility of T-reX is not limited to waste, it can be used to categorise and aggregate any technosphere exchange of interest (or customised grouping thereof), such as water, gas, and critical raw materials.

@@

The supplementary material in \autoref{sec:supplementary} contains metadata of
T-reX, along with a list of the constituent modules, a description of their
functions, and a detailed computational workflow. Further details can be found
in the GitHub repository and package documentation~\citep{mcdowall2024T-reXgithub, mcdowall2023T-reXdocs}.

@@

In the default configuration of T-reX, there are 10 waste categories which are
then further divided by their unit of measurement (kilograms and cubic meters)
to create a total of 20 waste methods. The waste categories include
aggregations such as `total waste' `hazardous waste' as well as, those for`'
waste sent to treatment by, for example, incineration or recycling. The full
details are listed in the supplementary material in~\autoref{sec:supplementary}
[section 3]. One advantage of T-reX is that it can identify `hidden' waste
exchanges that would otherwise be `consumed' by a treatment process and not
leave the technosphere.

In addition to these waste categories, T-reX can incorporate any number of
material demand categories, the defaults being based on the EU Critical Raw
Materials (CRM) list for 2023~\citep{eu2023crmstudy}, which contains 30
materials considered essential to the EU economy and also at risk of supply
disruption. Especially valuable in the case of CRMs is T-reX's ability to
uncover and quantify these threads of background consumption, which can be
inconspicuously embedded in products or consumed in seemingly distant supply
chain activities. The identity of the materials considered, and their
categorical groupings, are easily customisable by the user. The full list of 59
materials included in the default configuration is provided in the
supplementary material in~\autoref{sec:supplementary}.

@@

T-reX, in contrast, uses accounting for material demand and waste generation
based on exchanges solely within the technosphere, a material basis. This
offers a different perspective, allowing for the estimation of overall supply
chain material demands that consider the entire life cycle of an activity,
including non-direct effects on the market, such as a 'negative consumption' of
other materials due their co-production with the functional unit. Consider a
demand for an activity containing a metal, for example; while the existing
material use methods allow one to calculate the total mass of that metal that
is extracted from the environment, T-reX can provide insight into the broader
supply chain impacts of the demand for this metal. If the production of other
materials is attributed to the production of this metal, these would appear as
negative material demands in the T-reX results---supply chain pressure for one
material can result in lessening of supply chain pressure for another. For
example, in the results of the Li-ion battery case study , this is indeed the
case for nickel demand, which, in the final inventory because of such
co-production, is counter-intuitively negative (due to co-production and
substitution in the LCA system models) despite the presence of nickel in the
final products.

@@

An overview of the T-reX workflow is presented in
\autoref{fig:methods-flowchart}. Section four of the supplementary material in
\autoref{sec:supplementary} contains a more detailed computational flowchart.

\begin{figure*}
	\centering
	\caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}
	\includegraphics[width=1.7\columnwidth]{figures/T-reX_method.pdf}\label{fig:methods-flowchart}
\end{figure*}

@@

An example of the output from the application of T-reX has been included in the
supplementary material in section 6 of the ~\autoref{sec:supplementary}. The
manipulated \texttt{ecoinvent} databases (which are the main product of T-reX)
can be recreated by following the instructions in the package
documentation~\citep{mcdowall2023T-reXdocs}.

@@

As described in~\autoref{sec:method-casestudy}, this case study calculated the
waste and material footprints (with a variety of other indicators) for the
unaltered background inventories of five Li-ion batteries with the functional
unit being 1~kg of the battery at the global market. The purpose of this simple
case study was to test, verify, and demonstrate the functionality and
limitations of T-reX. This section includes some highlights of the results,
with full results included in the supplementary material
(\autoref{sec:supplementary}). Because the `pseudo-LCIA' methods created by
T-reX are integrated into the user's \texttt{brightway} project as if they were
standard LCIA methods, the footprint calculations can be performed in the
customary way. In the supplementary material, there are screenshots of selected
results obtained using the \texttt{ActivityBrowser} software, including
contribution analysis and a Sankey diagram that dis-aggregates the final
footprint result over the activities in the supply chain.

@@

\autoref{fig:sankey_waste} presents a heavily abridged Sankey diagram depicting the supply chain flows of `total solid waste' attributed to the NMC 811 Li-ion battery at the global market in the \texttt{ecoinvent} database version 3.9.1. As the supply chains are expansive, this visualisation has been greatly simplified to be legible for print, multiple versions are included in the supplementary material~\autoref{sec:supplementary}. As is evident in the figure, the largest contributors to the waste footprint are the extraction and processing of the raw materials, with the largest flows coming from the extraction of cobalt, lithium, and nickel. Due to the nature of LCA, the waste flows do not represent the `real' physical flows directly, but rather an abstraction thereof based on, in this case, economic allocation---the market price of the functional unit produced by the activity~\citep{guinee2004economicallocation}. If an unallocated database were available, the waste flows could be calculated with T-reX in the same way and would be more directly related to the physical flows of the activities.

\begin{figure}[H!]
	\centering
	\includegraphics[width=0.8\columnwidth]{figures/T-reX_NMC811_WasteTotalSolid.pdf}
	\caption{A greatly simplified Sankey diagram from \texttt{ActivityBrowser showing the total solid waste footprint flows for the
	 NMC 811 battery supply chain at the global market in the \texttt{ecoinvent} database version 3.9.1}label{fig:sankey_waste}
\end{figure}

@@

\begin{figure*}[H]
	\centering
	\includegraphics[width=/textwidth]{figures/T-reX-wastefootprint-totalANDcarbon.pdf}
	\caption{T-reX calculated supply chain inventories for the five Li-ion batteries in the case study, showing their `footprints' for: (a) total solid waste, and (b) carbon dioxide waste (produced from carbon capture and storage (CCS) only). The footprints were modelled with prospective LCA databases from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:waste_totalANDcarbon}
\end{figure*}

\subsubsection{Contribution of `top-processes' in the supply chain}%\label{sec:results-case_study-topprocesses}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{figures/T-reX-wastefootprint-processcontributions.pdf}
	\caption{Contribution by the `top-processes' to graphite demand inventory of the LiMn\(_2\)O\(_4\) battery from 2020 to 2100 under the SSP2 scenario using the baseline RCPs of the REMIND model}\label{fig:top_contribution}
\end{figure}

%!!!!

@@

\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{figures/T-reX-wastefootprint-sectorcontributions.pdf}
	\caption{Contribution of industrial sectors to the natural gas footprint of the LFP battery from 2020 to 2100 under the SSP2 scenario using the baseline RCP of the REMIND model}\label{fig:cpc_contribution}
\end{figure}

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.9\columnwidth]{figures/T-reX-coalANDedip.pdf}
	\caption{Comparison of the (a) `T-reX – Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 – coal no longterm'. In the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
\end{figure*}

@@

% Loading bibliography database
\bibliography{01_T-reX_manuscript.bib}

% insert supplementary information
%\clearpage
%#\includepdf{T-reX_supplementary-material/00_T-reX_supplementary-material_main.pdf}

>>>>

@@

\def\tsc#1{\csdef{#1}{\textsc{\lowercase{#1}}\xspace}}
\tsc{SCM}
\tsc{EL}
\tsc{SC}
\tsc{CFB}

%\usepackage[numbers]{natbib}
\usepackage[authoryear]{natbib}
% \usepackage[authoryear,longnamesfirst]{natbib}
%\usepackage{booktabs}
%\usepackage[tt=false]{libertine}
%\usepackage{xurl}

%%%
\begin{document}
	\let\WriteBookmarks\relax
	\def\floatpagepagefraction{1}
	\def\textpagefraction{.001}

@@

	% Short author
	\shortauthors{McDowall, S2. et~al.}

	% Main title of the paper
	\title [mode = title]{T-reX: Quantifying waste and material footprints in current and future Life Cycle Assessment (LCA) databases}

	% First author
	%
	% Options: Use if required
	% eg: \author[1,3]{Author Name}[type=editor,
	%       style=chinese,
	%       auid=000,
	%       bioid=1,
	%       prefix=Sir,
	%       orcid=0000-0000-0000-0000,
	%       facebook=<facebook id>,
	%       twitter=<twitter id>,
	%       linkedin=<linkedin id>,
	%       gplus=<gplus id>]

	\author[1]{Stewart Charles McDowall}[orcid=0000-0002-5688-1279]

	% Corresponding author indication
	\cormark[1]
	% Email id of the first author
	\ead{s.c.mcdowall@cml.leidenuniv.nl}
	% URL of the first author
	\ead[url]{scmcd.ch}
	% Credit authorship
	\credit{Methodology, Software, Validation, Formal analysis, Investigation, Data curation, Writing: original draft, Writing: review \& editing, Visualization}

	\affiliation[1]{
		organization={Institute of Environmental Sciences (CML), Leiden University},
		addressline={P.O. Box 9518},
		city={Leiden},
		postcode={2300RA},
		state={South Holland},
		country={The Netherlands}}

@@

	% Corresponding author text
	\cortext[1]{Corresponding author}

	% Here goes the abstract
	\begin{abstract}
		The quintessential principle of the circular economy is to minimise material consumption and waste generation. Thus, identifying and quantifying waste and material flows is critical.

		Life Cycle Assessment (LCA) is a powerful quantitative method for this, especially given its capacity to expose the details of an activity's entire life cycle, often guiding the implementation of circular principles to where they could be most effective.

		Pursuant to these goals the authors have developed `T-reX', an analytical tool for LCA, written in Python. T-reX extends on the existing computational ecosystem centered on \texttt{Brightway} and \texttt{Premise} and can facilitate the quantification of user-defined supply chain demands for activities in current and prospective scenarios. T-reX streamlines database manipulation for LCA practitioners, integrating methods to aggregate and analyse demand inventories that could expose hotspots of hidden risk.

		With a simple case study using lithium-ion batteries, we demonstrate some of T-reX's potential by detailing and exploring aspects of their waste and material footprints. Since such footprints are often linked to negative externalities, T-reX could support sustainable decision-making and contribute to the development of the circular economy by promoting the understanding of our material consumption and waste generation.
	\end{abstract}

	% Use if graphical abstract is present
	\begin{graphicalabstract}
		\includegraphics[width=\columnwidth]{grabs_T-reX_graphical-abstract.pdf}
		%\includegraphics[width=\columnwidth]{grabs_T-reX_graphical-abstract_MID-RES.png}
	\end{graphicalabstract}

	% Highlights
	\begin{highlights}
		\item T-reX, a new open-source program for waste \& material footprints in Life Cycle Assessment (LCA)
		\item Assesses supply risks by calculating demand inventories for critical materials
		\item Simplifies quantification of user-specified waste and material footprint categories
		\item Rapidly identifies hotspots of waste generation and material demand in supply	chains
		\item Presents a case study of five lithium-ion batteries to demonstrate T-reX's potential
	\end{highlights}

	% Keywords
	% Each keyword is seperated by \sep

	\begin{keywords}
		circular economy, waste footprint, material footprint, life cycle assessment, critical raw materials, supply chain
	\end{keywords}

	\maketitle

	% Main text
	\begin{table*}[!htbp]
		\centering
		\caption{List of abbreviations}\label{tab:abbreviations}
		\begin{tabular}{ll}
			\toprule
			CCS	              & Carbon Capture and Storage                                          \\
			CF                & Carbon Footprint                                                    \\
			EF                & Ecological Footprint                                                \\
			CML               & Centrum voor Milieuwetenschappen (Centre for Environmental Science) \\
			CRM               & Critical Raw Material                                               \\
			CPC               & Cooperative Patent Classification                                   \\
			CSI               & Crustal Scarcity Indicator                                          \\
			CSP               & Crustal Scarcity Potential                                          \\
			EDIP              & Environmental Design of Industrial Products                         \\
			EEIOA             & Environmentally Extended Input-Output Analysis                      \\
			EOL               & End-of-Life                                                         \\
			FOEN              & Swiss Federal Office for the Environment                            \\
			IAM               & Integrated Assessment Model                                         \\
			IE                & Industrial Ecology                                                  \\
			LCA               & Life Cycle Assessment                                               \\
			LCI               & Life Cycle Inventory                                                \\
			LCIA              & Life Cycle Impact Assessment                                        \\
			LFP               & Lithium Iron Phosphate                                              \\
			Li-ion            & Lithium-ion                                                         \\
			LiMn\(_2\)O\(_4\) & Lithium Manganese Oxide                                             \\
			MF                & Material Footprint                                                  \\
			MFA               & Material Flow Analysis                                              \\
			NCA               & Nickel Cobalt Aluminum                                              \\
			NMC               & Nickel Manganese Cobalt                                             \\
			PWF               & Product Waste Footprint                                             \\
			RCP               & Representative Concentration Pathway                                \\
			ReCiPe            & A standard LCIA method set                                          \\
			SDG               & Sustainable Development Goals                                       \\
			SSPs              & Shared Socioeconomic Pathways                                       \\
			T-reX             & The Tool for analysing re-X in LCA                                  \\
			UBP               & Umweltbelastungspünkte (Environmental Impact Points)                \\
			UN                & United Nations                                                      \\
			WF                & Waste Footprint                                                     \\
			\bottomrule
		\end{tabular}
	\end{table*}

	\section{Introduction}\label{sec:introduction}  
	\subsection{Introduction to T-reX}\label{sec:intro-trex}

	The development of a circular economy has become a central focus in the pursuit of sustainability objectives toward curtailment of our environmental footprint within planetary boundaries~\citep{eu2019greendeal,eu2020circ,nl2023ceplan,nl2016ceplan,pardo2018ce,ellenmacarthur2015ce}.

	Fundamental to this development is a decrease in primary material consumption and a reduction of life cycle waste through the implementation of `re-X'v strategies (e.g., refuse, rethink, design for---and implementation of---repair, remanufacturing and recycling)~\citep{reike2018rex, eu2022ecodesign, eu2022repair, eu2015reman}. In addition to circular economy goals, contemporary geopolitical tensions in an ever more globalised world have highlighted the vulnerability of many advanced economies to intentional supply disruptions---wrought as an act of competition or even outrighthostility~\citep{jrc2023supplychain,hartley2024cepolitics,berry2023crm}.

	The concept of the `footprint' as an environmental sustainability indicator began with the Ecological Footprint (EF)~\citep{wackernagel1994ecologicalfootprint} and after being popularised by the Carbon Footprint (CF)~\citep{cucek2015environmentalfootprints} the `footprint family' has adopted many additional metrics---albeit without yet
	coalescing into a coherent or consistent framework~\citep{giampietro2014footprintstonowhere,vanham2019footprints,ridoutt2013footprints}. More recently, the footprint
	collection has been extended to include the Material Footprint (MF)~\citep{weidmann2013materialfootprint}, which is often encountered in the
	`macro methods' of Industrial Ecology (IE), such as environmentally Extended Input-Output Analysis (EEIOA)~\citep{lenzen2022materialfootprint} and Material
	Flow Analysis (MFA)~\citep{schaffartzik2013mfafootprint}. Whether at the level of products, entire industries, nations, or even continents, the material
	footprint aims to quantify the total material consumed in supply chains. It has been shown that the MF can be `highly representative of damage to human health
	and biodiversity'~\citep{steinmann2017resourcefootprints} and indeed, this metric was recently beatified by the United Nations (UN), becoming the `core
	official indicator' for targets 8.4 and 12.2 of the Sustainable Development Goals (SDGs)~\citep{lenzen2022materialfootprint}.

@@

	WF and MF, the two metrics of focus in the study, can provide a comprehensive assessment of potential environmental impacts across the supply chain, encapsulating both resource use and pollution/waste generation, while offering insights at diverse scales, from individual activities to global systems,facilitating communication with a diverse range of stakeholders. Thus, to reduce the negative externalities of human consumption and improve supply chain resilience, it is essential to uncover, disaggregate, and quantify the material and waste footprints of human activities in as much detail as possible~\citep{bisinella2024wastelca,towa2020wastefootprint,berger2020mineralsinlca,sonderegger2020mineralsinlca}.

@@

	To facilitate the quantification of waste and material flows in LCA, we
	developed a Python package to extend the \texttt{brightway} open-source LCA
	framework~\citep{mutel2017brightway} and track these exchanges by translating
	them into inventory indicators and `pseudo' LCA impact (LCIA) categories. T-reX
	is `T[ool for] re-X' (reduce, reuse, recycle, etc.) and enables LCA
	practitioners to manipulate their databases to allow them to easily aggregate
	the mass and volume of any desired exchange, and to create flexible categories
	that differentiate between material categories, waste types, and treatment or
	EOL handling.

@@

	~\cite{laurenti2023wastefootprint} developed a method to calculate the waste footprint of a product or service based on solving the demand vectors of the activities, while also proposing simple measures to quantify waste `hazardousness' and `circularity' with this data. The method presented, as the authors explain, suffers from errors due to double counting due to the way that exchanges are identified and `tagged'. In T-reX, the source database is deconstructed into a list of separate exchanges, and only those explicitly defined as hazardous are marked as such. The `T-reX' Python software tool presented herein provides a more flexible, transparent, and user-friendly approach to quantifying waste flows in LCA\@. Moreover, the utility of T-reX is not limited to waste, it can be used to categorise and aggregate any technosphere exchange of interest (or customised grouping thereof), such as water, gas, and critical raw materials.

@@

	The supplementary material in \autoref{sec:supplementary-1} contains metadata of
	T-reX, along with a list of the constituent modules, a description of their
	functions, and a detailed computational workflow. Further details can be found
	in the GitHub repository and in the package documentation~\citep{mcdowall2024T-reXgithub, mcdowall2023T-reXdocs}.

@@

	In the default configuration of T-reX, there are 10 waste categories which are
	then further divided by their unit of measurement (kilograms and cubic meters)
	to create a total of 20 waste methods. The waste categories include
	aggregations such as `total waste' `hazardous waste' as well as, those for
	`waste sent to treatment' by, for example, incineration or recycling. The full
	details are listed in the supplementary material in~\autoref{sec:supplementary-3}. One advantage of T-reX is that it can identify `hidden' waste
	exchanges that would otherwise be `consumed' by a treatment process and not
	leave the technosphere.

	In addition to these waste categories, T-reX can incorporate any number of
	material demand categories, the defaults being based on the EU Critical Raw
	Materials (CRM) list for 2023~\citep{eu2023crmstudy}, which contains 30
	materials considered essential to the EU economy and also at risk of supply
	disruption. Especially valuable in the case of CRMs is T-reX's ability to
	uncover and quantify these threads of background consumption, which can be
	inconspicuously embedded in products or consumed in seemingly distant supply
	chain activities. The identity of the materials considered, and their
	categorical groupings, are easily customisable by the user. The full list of 59
	materials included in the default configuration is provided in the
	supplementary material in~\autoref{sec:supplementary-3}.

@@

	T-reX, in contrast, uses accounting for material demand and waste generation
	based on exchanges solely within the technosphere, a material basis. This
	offers a different perspective, allowing for the estimation of overall supply
	chain material demands that consider the entire life cycle of an activity,
	including non-direct effects on the market, such as a `negative consumption' of
	other materials due their co-production with the functional unit. Consider a
	demand for an activity containing a metal, for example; while the existing
	material use methods allow one to calculate the total mass of that metal that
	is extracted from the environment, T-reX can provide insight into the broader
	supply chain impacts of the demand for this metal. If the production of other
	materials is attributed to the production of this metal, these would appear as
	negative material demands in the T-reX results---supply chain pressure for one
	material can result in lessening of supply chain pressure for another. For
	example, in the results of the Li-ion battery case study, this is indeed the
	case for nickel demand, which, in the final inventory because of such
	co-production, is counter-intuitively negative (due to co-production and
	substitution in the LCA system models) despite the presence of nickel in the
	final products.

@@

	An overview of the T-reX workflow is presented in
	\autoref{fig:methods-flowchart}. The supplementary material in
	\autoref{sec:supplementary-3} provides more detailed information.

	\begin{figure*}[]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX_method.pdf}
		\caption{The workflow of T-reX presented in a generalised (a) and computational (b) format. Application of T-reX to one or more LCA databases creates a \texttt{brightway} project with customised methods and activities that can be used to calculate waste and material inventory footprints of the activity's supply chain.}\label{fig:methods-flowchart}
	\end{figure*}

@@

	An example of the output from the application of T-reX has been included in the
	supplementary material~\autoref{sec:supplementary-5}. The
	manipulated \texttt{ecoinvent} databases (which are the main product of T-reX)
	can be recreated by following the instructions in the package
	documentation~\citep{mcdowall2023T-reXdocs}.

@@

	As described in~\autoref{sec:method-casestudy}, this case study calculated the
	waste and material footprints (with a variety of other indicators) for the
	unaltered background inventories of five Li-ion batteries with the functional
	unit being 1~kg of the battery at the global market. The purpose of this simple
	case study was to test, verify, and demonstrate the functionality and
	limitations of T-reX. This section includes some highlights of the results,
	with full results included in the supplementary material
	(\autoref{sec:supplementary-7} and \autoref{sec:supplementary-8}). Because the `pseudo-LCIA' methods created by
	T-reX are integrated into the user's \texttt{brightway} project as if they were
	standard LCIA methods, the footprint calculations can be performed in the
	customary way. In the supplementary material, there are screenshots of selected
	results obtained using the \texttt{ActivityBrowser} software, including
	contribution analysis and a Sankey diagram that dis-aggregates the final
	footprint result over the activities in the supply chain.

@@

	\autoref{fig:sankey_waste} presents a heavily abridged Sankey diagram depicting the supply chain flows of `total solid waste' attributed to the NMC 811 Li-ion battery at the global market in the \texttt{ecoinvent} database version 3.9.1. As the supply chains are expansive, this visualisation has been greatly simplified to be legible for print, multiple versions are included in the supplementary material in~\autoref{sec:supplementary-8}. As is evident in the figure, the largest contributors to the waste footprint are the extraction and processing of the raw materials, with the largest flows coming from the extraction of cobalt, lithium, and nickel. Due to the nature of LCA, the waste flows do not represent the `real' physical flows directly, but rather an abstraction thereof based on, in this case, economic allocation---the market price of the functional unit produced by the activity~\citep{guinee2004economicallocation}. If an unallocated database were available, the waste flows could be calculated with T-reX in the same way and would be more directly related to the physical flows of the activities.

	\begin{figure}[!htbp]
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX_NMC811_WasteTotalSolid.pdf}
		\caption{A greatly simplified Sankey diagram from \texttt{ActivityBrowser} showing the total solid waste footprint flows for the
			NMC 811 battery supply chain at the global market in the \texttt{ecoinvent} database version 3.9.1}\label{fig:sankey_waste}
	\end{figure}

@@

	\begin{figure*}[!htbp]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX-wastefootprint-totalANDcarbon.pdf}
		\caption{T-reX calculated supply chain inventories for the five Li-ion batteries in the case study, showing their `footprints' for: (a) total solid waste, and (b) carbon dioxide waste (produced from carbon capture and storage (CCS) only). The footprints were modelled with prospective LCA databases from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model.}\label{fig:waste_totalANDcarbon}
	\end{figure*}

	\subsubsection{Contribution of `top-processes' in the supply chain}\label{sec:results-case_study-topprocesses}

@@

	\begin{figure}[!htbp]
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX-wastefootprint-processcontributions.pdf}
		\caption{Contribution by the `top-processes' to graphite demand inventory of the LiMn\(_2\)O\(_4\) battery from 2020 to 2100 under the SSP2 scenario using the baseline RCPs of the REMIND model}\label{fig:top_contribution}
	\end{figure}

@@

	\begin{figure}
		\centering
		\includegraphics[width=0.8\columnwidth]{T-reX-wastefootprint-sectorcontributions.pdf}
		\caption{Contribution of industrial sectors to the natural gas footprint of the LFP battery from 2020 to 2100 under the SSP2 scenario using the baseline RCP of the REMIND model}\label{fig:cpc_contribution}
	\end{figure}

@@

	\begin{figure*}[!htbp]
		\centering
		\includegraphics[width=1.8\columnwidth]{T-reX-coalANDedip.pdf}
		\caption{Comparison of the (a) `T-reX – Coal (black)' demand footprint with the (b) conventional LCIA method `EDIP 2003 – coal no longterm'. In the case study from 2020 to 2100 under the SSP2 scenario using the baseline and PkBudg500 RCPs of the REMIND model for all five batteries.}\label{fig:comparison_methods}
	\end{figure*}

@@

	% Loading bibliography database
	\bibliography{01_T-reX_manuscript}
